The brightness and spatial distributions of terrestrial radio sources

### 2013-01-15 articles/Article-RFI-distribution/article-rfi-distribution.tex
Correcting Iliev''s name
Some fixes based on comments by internal ref

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{2,3}$,
S.~Zaroubi$^2$,
W.~Brouw$^{2,3}$,
B.~Ciardi$^4$,
I.~T.~Iliav$^5$,
\newauthor
G.~J.~A.~Harker$^6$,
L.~V.~E.~Koopmans$^2$,
G.~Mellema$^7$,
S.~J.~Wijnholds$^3$
\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. First we will derive the expected intrinsic source distribution. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

@@

Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$. For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal. The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side (e.g., because of the curvature of the Earth). Solving this equality, one finds that
\begin{equation} \label{eq:lower-limit-1}
S_L = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta}\left(N_\textrm{part} - N_\textrm{total} \right) + S_1^{\alpha+1}}.
\end{equation}

@@

% TODO appendix with calculation of 2D case == 3D case?

>>>>

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{2,3}$,
S.~Zaroubi$^2$,
W.~Brouw$^{2,3}$,
B.~Ciardi$^4$,
I.~T.~Iliev$^5$,
\newauthor
G.~J.~A.~Harker$^6$,
L.~V.~E.~Koopmans$^2$,
G.~Mellema$^7$,
S.~J.~Wijnholds$^3$
\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}). For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources. However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.

First we will derive the expected intrinsic source distribution for interfering radio sources. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

\subsection{Spherical case} \label{sec:spherical-geometry}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna. As shown in Appendix~\ref{sec:proof-spherical-case}, Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, has little effect on the $\log N$ -- $\log S$ histograms. Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, $I$ might follow a power law by itself. Strong transmitters are probably on average transmitting longer, thus will also contaminate more samples. This weakens the effect of $I$ on the histogram somewhat. %TODO

@@

Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$. For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal. The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side, e.g., because of the curvature of the Earth. Solving $N_a = N_\textrm{total}-N_\textrm{part}$ results in
\begin{equation} \label{eq:lower-limit-1}
S_L = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta}\left(N_\textrm{part} - N_\textrm{total} \right) + S_1^{\alpha+1}}.
\end{equation}

@@

\appendix
\section{Calculations with spherical geometry} \label{sec:proof-spherical-case}
As stated in \S\ref{sec:spherical-geometry}, the distance distribution of sources for a flat geometry and a spherical surface are equal for distances smaller than the diameter of the Earth. We proof this by showing that the spherical distance function $\tilde{F}_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-spherical} equals the flat distance function $F_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-distance-function}.
\begin{proof}
\begin{eqnarray} \notag
\tilde{F}_{\textrm{distance} \le r}(r) & = & \rho 2 \pi R^2 \left[1  - \cos\left(2 \arcsin \frac{r}{2R}\right)\right] \\\notag
& & \textit{\{\hspace{1mm}Using $\cos 2\phi = 1 - 2\sin^2 \phi$\hspace{1mm}\}} \\\notag
& = & \rho 2 \pi R^2 \left[1 - \left(1 - 2 \sin^2(\arcsin \frac{r}{2R})\right)\right] \\\notag
& = & \rho 2 \pi R^2 \left[2 \sin^2(\arcsin \frac{r}{2R})\right] \\\notag
& & \textit{\{\hspace{1mm}Assuming $r \le 2R$\hspace{1mm}\}} \\\notag
& = & \rho 2 \pi R^2 \left[2 \left(\frac{r}{2R}\right)^2\right] \\\notag
& = & \rho \pi r^2 \\
& = & F_{\textrm{distance} \le r}(r)
\end{eqnarray}
\end{proof}
